Make any stupid plot you want

### 2021-04-29 draft/sections/intro.tex
data, closes out background

Visualizations assume the structure of the input data, as described in Tory and Möller's taxonomy \cite{toryRethinkingVisualizationHighlevel2004}, which leads to many building block libraries implementing multiple models of data to support these different visualizations. For example, in Matplotlib, D3\cite{bostockDataDrivenDocuments2011} VTK \cite{hanwellVisualizationToolkitVTK2015,geveciVTK2012} and MayaVi\cite{RamachandranMayaVI2011}, every plot is defined in terms of the \textit{continuity} of the data it expects as input. VTK has explicitly codified this in terms of \textit{continuity} based data representations, as illustrated in figure~\ref{fig:intro:data:format}. Downstream library developers impose some coherency by writing domain specific libraries with assumed data structures on top of the building block libraries--for example Seaborn\cite{waskom2020seaborn} and Titan\cite{brianwylieUnifiedToolkitInformation2009} assume a relational database, xarray\cite{hoyer2017xarray} and ParaView\cite{ahrens2005paraview} assume a data cube--but must work around the incoherencies in the building block libraries to do so. Our model navigates the tradeoff between coherency and extensibility by proposing functional composable well constrained visual components that take as input a structure aware data abstraction general enough to provide a common interface for many different types of visualizations.

\subsection{Data}
\label{sec:intro_data}
One such general abstraction are fiber bundles, which Butler proposed as a core data structure for visualization because they encode data continuity separately from the variable properties and are flexible enough to support discrete and ND continuous datasets \cite{butlerVisualizationModelBased1989,butlerVectorBundleClassesForm1992}. 
Since Butler's model lacks a robust way of describing variables, we can encode a schema like description of the data in the fiber bundle by folding in Spivak's topological description of data types \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES}. In this work we will refer to the points of the dataset as \textit{records} to indicate that a point can be a vector of heterogenous elements. Each \textit{component} of the record is a single object, such as a temperature measurement, a color value, or an image. We also generalize \textit{component} to mean all objects in the dataset of a given type, such as all temperatures or colors or images. The way in which these records are connected is the \textit{connectivity}, \textit{continuity}, or more generally \textit{topology}.

@@

The \textit{continuity} can often be described by some variables in the dataset; this is formalized Munzner's notion of metadata as \textit{keys} into the data structure that return associated \textit{values}\cite{munznerChDataAbstraction}. As shown in \autoref{fig:intro:keys:values}, keys can be labeled indexes, such as the attribute name and row ID, or physical entities such as locations on a map. We propose that information rich metadata are part of the components and instead the values are keyed on coordinate free structural ids. In contrast to Munzner's model where the semantic meaning of the key is tightly coupled to the position of the value in the dataset, our model considers keys to be a pure reference to topology. This allows the metadata to be altered, without imposing new semantics on the underlying structure, for example by changing the coordinate systems or time resolution. This value agnostic model also supports encoding datasets where there may be multiple independent variables - such as a measure of plant growth given variations in water, sunlight, and time - without having to assume any one variable is inducing the change in growth. For building block library developers, this means the components are able to fully traverse the data structures without having to know anything about the values or the semantic meaning of the structure. Since these components are by design \textit{equivariant} and \textit{continuity} preserving, domain specific library developers in different domains that both rely on the same continuity, for example 2D continuity, can then safely reuse the components to build tools that can safely make domain specific assumptions.

\subsection{Contribution}
The contribution of this work is 
\begin{enumerate}
  \item formalization of the topology preserving relationship between data and graphic via continuous maps \autoref{sec:math:graphic:base}
  \item formalization of property preservation from data component to visual representation as monoid action equivariant maps \autoref{sec:math:artist:nu}
  \item functional oriented visualization architecture built on the mathematical model to demonstrate the utility of the model \autoref{sec:math:artist:q}
  \item prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model. \autoref{sec:math:code:artist}
\end{enumerate}
\end{document}

>>>>

Visualizations assume the structure of the input data, as described in Tory and Möller's taxonomy \cite{toryRethinkingVisualizationHighlevel2004}, which leads to many building block libraries implementing multiple models of data to support these different visualizations. For example, in Matplotlib, D3\cite{bostockDataDrivenDocuments2011} VTK \cite{hanwellVisualizationToolkitVTK2015,geveciVTK2012} and MayaVi\cite{RamachandranMayaVI2011}, every plot is defined in terms of the \textit{continuity} of the data it expects as input. VTK has explicitly codified this in terms of \textit{continuity} based data representations, as illustrated in figure~\ref{fig:intro:data:format}. Downstream library developers impose some coherency by writing domain specific libraries with assumed data structures on top of the building block libraries--for example Seaborn\cite{waskom2020seaborn} and Titan\cite{brianwylieUnifiedToolkitInformation2009} assume a relational database, xarray\cite{hoyer2017xarray} and ParaView\cite{ahrens2005paraview} assume a data cube--but must work around the incoherencies in the building block libraries to do so. Our model navigates the tradeoff between coherency and extensibility by proposing functional composable well constrained visual components that take as input a structure aware data abstraction general enough to provide a common interface for many different types of data continuities.

\subsection{Data}
\label{sec:intro_data}
Fiber bundles were proposed by Butler as a core data structure for visualization because they encode data \textit{continuity} separately from the components of the dataset\cite{butlerVisualizationModelBased1989,butlerVectorBundleClassesForm1992}. Butler's model lacks a robust way of describing variables; therefore we encode a schema like description of the data in the fiber bundle using Spivak's topological description of data types \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES}. In this work, we refer to the points of the dataset as \textit{records}, as defined by Spivak. Each \textit{component} of the record is a single object, such as a precipitation measurement, a station name, or an image. We generalize \textit{component} to mean all objects in the dataset of a given type, such as all precipitation values or station names or images. The way in which these records are connected is the \textit{continuity} or more generally topology.

@@

The \textit{continuity} can be described in some datasets by components of the dataset. This is formalized by Munzner's notion of metadata as \textit{keys} into the data structure that return associated \textit{values}\cite{munznerChDataAbstraction}. As shown in \autoref{fig:intro:keys:values}, keys can be labeled indexes, such as the attribute name and row ID, or semantically significant physical entities such as locations on a map. In contrast to Munzner's model, in out model components may describe the keys but are never themselves the keys; instead we propose that keys are points on a topological space encoding the continuity of the data. This allows the metadata to be altered without imposing new semantics on the underlying structure, for example by changing the coordinate systems or time resolution. This value agnostic model also supports encoding datasets where there may be multiple independent variables without having to assume any one variable is inducing the change , for example measures of plant growth given variations in water, sunlight, and time.

For building block library developers, a non-semantic model of data continuity allows for the implementation of components that can traverse data structures without having to know the semantics of the data. Since these building block components are by design \textit{equivariant} and \textit{continuity preserving}, domain specific library developers in different domains that rely on the same continuity, for example 2D continuity, can use the same components to build tools that can make domain specific assumptions. One way we evaluate our model is by reproducing fully constrained domain specific library functionality using our proposed architecture.

\subsection{Contribution}
The contribution of this work is 
\begin{enumerate}
  \item formalization of the topology preserving relationship between data and graphic via continuous maps \autoref{sec:math:graphic:base}
  \item formalization of property preservation from data component to visual representation as monoid action equivariant maps \autoref{sec:math:artist:nu}
  \item functional oriented visualization architecture built on the mathematical model to demonstrate the utility of the model \autoref{sec:math:artist:q}
  \item prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model. \autoref{sec:code:artists}
\end{enumerate}
\end{document}
